Lowestcase and Uppestcase letters: Advances in Derp Learning

### 2021-03-14 lowercase/paper/lowercase.tex
Fix bug in remapping that meant it didn't affect most of the image, ugh. Some additions to paper

\title{Lowestcase and Uppestcase letters: Further adventures in Derp Learning}
\author{Dr.~Tom~Murphy~VII~Ph.D.}\thanks{
Copyright \copyright\ 2021 the Regents of the Wikiplia Foundation.
Appears in SIGBOVIK~2021 with the
OS2TypoLinegap of the Association for Computational Heresy; {\em IEEEEEE!}
press, Verlag-Verlag volume no.~0x40-2A. 1 em}

@@

\renewcommand\paragraph[1]{\smallskip{\bf #1}\enspace}

@@

I computed some stats on the font database, and determined the
following parameters for the fixed-size SDFs we train on. The images
are $36 \times 36$ pixels. The character box is placed such that there
are $2$ pixels of top padding, and $9$ pixels of left and bottom
padding. The character box is only ``nominal'' in the sense that the
font's contours can exceed its bounds, and this is completely normal
for a letter like \letterform{j} (which goes below the baseline and
often hangs to the left of the origin as well). I used an ``on-edge
value'' of 220 (because much more of the SDF is outside the letter
than inside) and the distance is scaled as 15 units per pixel (so that
pixels on the outer edge often have non-zero values). Compared to the
first version, I was somewhat more permissive in what fonts I trained
on, since there was no inherent limit to the number of contours or their
complexity. I did exclude fonts whose rasterizations exceeded the
bounds of the SDF, which is possible (very wide \letterform{W} or
low-descending \letterform{j} perhaps) but rare.

@@

So I did that and let it run for a month. Actually I had to start over
several times with different parameters and initialization weights
because it would get stuck (Figure~\ref{fig:nans}) right away. I
prayed to the dark wizard of hyperparameter tuning until he smiled
upon my initial conditions, knowing that somewhere he was adding
another tick-mark next to my name in a tidy but ultimately scary
Moleskine notebook that he bought on a whim in the Norman Y.~Mineta
San Jose International Airport on a business trip, and still feels
was overpriced for what it is.

@@

The training error over time appears in
Figure~\ref{fig:sdfmergederror}. It looks like the ones I have seen in
machine learning papers, although I don't like to read other people's
papers because it just seems like spoilers.

@@

@@

\label{sec:fiddly}

% TODO loss over time

% some other things that are worth a mention:
%   autoparallel
%   efficiency improvements (restrict, compile-time constants)

@@

% stretch the loss function near the onedge value

@@

% lowercase uppercase letters etc.

>>>>

\title{Lowestcase and Uppestcase letters: Advances in Derp Learning}
\author{Dr.~Tom~Murphy~VII~Ph.D.}\thanks{
Copyright \copyright\ 2021 the Regents of the Wikiplia Foundation.
Appears in SIGBOVIK~2021 with the
OS2TypoLinegap of the Association for Computational Heresy; {\em IEEEEEE!}
press, Verlag-Verlag volume no.~0x40-2A. 1 em}

@@

\renewcommand\paragraph[1]{\smallskip \noindent{\bf #1}\enspace}

@@

I computed some stats on the font database, and determined the
following parameters for the fixed-size SDFs we train on. The images
are $36 \times 36$ pixels. The character box is placed such that there
are $2$ pixels of top padding, and $9$ pixels of left and bottom
padding. The character box is only ``nominal'' in the sense that the
font's contours can exceed its bounds, and this is completely normal
for a letter like \letterform{j} (which goes below the baseline and
often hangs to the left of the origin as well). I used an ``on-edge
value'' of 0.862 (because much more of the SDF is outside the letter
than inside) and the distance is scaled as 0.059 units per pixel (so
that pixels on the outer edge often have non-zero values). Compared to
the first version, I was somewhat more permissive in what fonts I
trained on, since there was no inherent limit to the number of
contours or their complexity. I did exclude fonts whose rasterizations
exceeded the bounds of the SDF, which is possible (very wide
\letterform{W} or low-descending \letterform{j} perhaps) but rare.

@@

So I did that and let it run for a month. Actually I had to start over
several times with different parameters and initialization weights
because it would get stuck (Figure~\ref{fig:nans}) right away or as
soon as I looked away from the computer. I prayed to the dark wizard
of hyperparameter tuning until he smiled upon my initial conditions,
knowing that somewhere he was adding another tick-mark next to my name
in a tidy but ultimately terrifying Moleskine notebook that he bought
on a whim in the Norman Y.~Mineta San Jose International Airport on a
business trip, and still feels was overpriced for what it is.

@@

\subsection{Fiddly bits} \label{sec:fiddly}
The training error over time appears in
Figure~\ref{fig:sdfmergederror}. It looks like the ones I have seen in
machine learning papers, although I don't like to read other people's
papers because it just seems like spoilers. There are several
noticeable events in the curve, which came from me fiddling with the
parameters or network as it ran. Here are some of the things I did:

\paragraph{Vacuuming and culling.} Sometimes a node will just be
dead (basically never activates) or an edge weight will be nearly
zero. In these cases an equivalent, tidier network can be made by
dropping the node or edge. Periodically I would perform these
processes, sometimes feeling particularly choppy and removing like
10\% of the parameters at a time. If these parameters are truly
useless with no hope of recovery, we simply get faster training
because there's less work to do.

\paragraph{Widening.} The opposite thing is to introduce new nodes.
Adding nodes to hidden layers is pretty easy. The thing that worked
best for me is to increase the size of the layer by 10--15\%, where
each new node has random incoming weights and bias 0. Then for each
node on the next layer, I add edges to some subset of these new nodes
(again generally 10\% of them) with weight 0. Since this weight is
zero, the network computes the same function, but has new gradients
to explore (in practice, it then experiences some shock after a
few training rounds, but then quickly fine-tunes this away). More
parameters means slower training, but also more potential to learn
interesting functions, or overfit!

\paragraph{Deepening.} It's also possible to add layers once the
network is trained. This can be done anywhere, but I liked doing it on
the output layer because this gets the most direct feedback from the
training examples, and so it updates quickly and changes there are
easy to understand. If you append the identity matrix (new layer is
the same size as the previous; each node has weight 1.0 to its
corresponding node and 0.0 elsewhere) then this network computes
the same function but has new gradients to explore. Adding a layer
did seem to help unlock a new training regime
(Figure~\ref{fig:sdfmergederror}); subjectively it also reduced
some weird artifacts in the SDFs that the model used to predict (makes
sense; this most natural thing for this layer to do is learn how to
predict a ``correction'' from the old prediction, for example
by smoothing/sharpening it). This seems to be borne out by the
weights, which are also fun to look at (Figure~\ref{fig:lastlayer}).

\paragraph{Generating features.} On the other side, randomly sampling
pixels from the input SDF does work, but I superstitiously believed
that it might be better to have more spatially meaningful features. I
wrote a program to generate a bunch of random simple features (one
line/blob with positive weights, one line/blob with negative weights).
It then chooses a set of them that are both {\em good} (maximum
standard deviation on a sample of training data) and {\em different
  from one another} (redundant or even partially redundant features
are less valuable). It was nice to satisfy my supersition, and the
dark wizard of superstitious fiddling with neural networks in the
hope that they do the thing was pleased as well. The features are
at least handsome (Figure~\ref{fig:makefeatures}).

I assume these are all standard things that neural people do, but they
do better and smarter versions of them because they are willing to
read other people's papers instead of trying to figure things out from
scratch all the time. But you gotta occupy yourself somehow while it
crunches for a month.

\begin{figure}[ht]
\centering
  \includegraphics[width=0.99 \linewidth]{makefeatures}
  \caption{ Some randomly-generated features with the selected ones
    outlined in magenta, mostly shown here for aesthetic reasons.
    Savvy Twitter user {\tt @iotathisworld} sees this as ``the classic
    question: machine vision classification or 90s roller rink carpet
    pattern?'' to which I deflect: ``Sorry, it's actually modern day
    Port Authority bus upholstery or Gram stain of same!'' (But
    actually machine vision classification was basically correct.)
  } \label{fig:makefeatures}
\end{figure}

@@

\begin{figure}[ht]
\centering
  \includegraphics[width=0.99 \linewidth]{lastlayer}
  \caption{ The bottom-right corner of the weight matrix for the final
    layer of the network. This layer was added to the network after
    1.36 million rounds, initially as the identity matrix, and so can
    be thought of partly as a correction of the network's output prior
    to that round (though that part of the network continues to
    evolve). The x-axis is the output nodes, and the y-axis is the
    nodes of the previous layer. Note for example that the last 26
    columns look pretty different; these are the predictors for the 26
    letters, which occur in the output after the SDF pixels. Green
    means positive and red means negative, so if you are looking at
    this in a black-and-white printout, that may explain your current
    confusion. The exact diagonal is a strong green, close to 1.0,
    although over time these weights do diverge from the identity
    somewhat. In the bottom-right corner of size $26^2$, we are
    looking at how the 26 letter predictors are derived from the
    previous layer's predictions. We see that most letters are
    negatively correlated (makes sense; only one will ever be 1.0)
    although there are some oddities (probably because it found some
    other, better correlates). Nodes on this new layer have dense
    references to these 26 predictions on the previous; this means
    that the bottom row kind of represents biases for each of the 26
    letters (what does an average 'e' look like?). I also included a
    dense region above that, but this appears to have simply evolved
    the same way as other $36^2$ chunks have (the rest are sparse).
    These chunks have a large amount of spatial similarity (suggesting
    that the sparse sampling would be adequate), with a meaning like
    ``if this area of the image is bright, then this pixel should
    be less bright.'' It is interesting that the pixels immediately
    next to the diagonal are almost always strongly negative
    (sharpening operation). ``Thank you for attending my TED talk.''
    ---\,Figure~\ref{fig:lastlayer}}
  \label{fig:lastlayer}
\end{figure}

For completeness, some other innovations that I feel are worth
mentioning:

Making the GPU code faster has really high value (could save weeks of
waiting). Since I am using OpenCL (whoa, yeah, stop me right there, I
know) I found a good trick was to generate different OpenCL code with
constants baked for each layer (for example their size and
\verb+indices_per_node+); this allows some of the inner loops to use
faster tricks for e.g.~multiplication by a compile-time constant
instead of depending on an argument or data. I have different routines
for sparse and dense layers. It might even make sense to recompile the
kernels for other parameters that change over the lifetime of
training, like the learning rate. The {\tt fma} instruction (so named
for the physical law $F=MA$) is a bit faster than
\verb|potential += w * v|, and I guess the compiler can't do this
itself because of IEEE horrors. But like, who cares? In my opinion you
should be able to put it in ``fast machine learning mode'' where it
readily makes precision errors, with a command-line option like
\verb+--fml+. With all the tweaking, the easiest win was to use the
{\tt restrict} keyword on arguments to tell the compiler that the
input cannot alias the output, for example; this presumably helps
it schedule instructions better.

Various things in training run in parallel threads (e.g.~processing
fonts, but also moving data to the GPU, backpropagation for each
example, etc.). For a long time I had just been explicitly setting
parallelism using superstitious constants. For this project I finally
just wrote something that would automatically and empirically
determine the number of threads that yielded the highest throughput,
and persisted that information across program starts. This was a
good idea and enters the permanent rotation.

The actual error on the predicted SDFs is pretty low; for the
\makelowercase\ model it is around 31.3, which is like if 2.4\% of the
pixels were (completely) wrong, but the rest is exactly correct. In
reality, of course, the error is distributed throughout the pixels,
and some errors are a lot more important than others. Particularly,
near the threshold value, a pixel goes from from being considered ``in
the letter'' to ``outside'' with tiny changes in its value. Changes to
a pixel with a value near 0.0 or 1.0 usually doesn't affect the output
shape at all, in contrast. So one thing I did was map the loss
function (comparing expected pixel value to actual) to ``stretch out''
the region near the threshold, increasing the penalty (basically, the
derivative) in that region and decreasing it elsewhere. Looking at the
code again right now, I realize that I only applied this to the first
row of the SDF (\verb+idx < SDF_SIZE+ instead of
\verb+SDF_SIZE * SDF_SIZE+), so that was dumb. I will say in my
defense that at least I felt disappointed that it didn't seem to make
a difference!

@@

@@

\subsection{Upperercase and Lowerercase fonts} \label{sec:fonts}

Let's make some fonts!

% autotracing in here, I think?

@@

% lowercase uppercase letters etc.
% iterate this a lot of times?

% run it to infinity to see the uppestcase, etc.
% if these vanish, then we have a fixed point.
% if not, or perhaps by doing some auto-gamma, then
% it's like a fractal zoomer, and we can set the
% record on the "most uppercase letter" by running
% a jazillion times.
